---
title: "Lab Assignment 4: Spatial Predictive Analysis"
subtitle: "Spatial Predictive Analysis of Graffiti Removal"
author: "Ming Cao"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

In this lab, you will apply the spatial predictive modeling techniques demonstrated in the class exercise to a 311 service request type of your choice. You will build a complete spatial predictive model, document your process, and interpret your results.

### Timeline & Deliverables

**Due Date:** November 17, 2025, 10:00AM

**Deliverable:** One rendered document, posted to your portfolio website.

### Learning Goals

By completing this assignment, you will demonstrate your ability to:

- Adapt example code to analyze a new dataset
- Build spatial features for predictive modeling
- Apply count regression techniques to spatial data
- Implement spatial cross-validation
- Interpret and communicate model results
- Critically evaluate model performance

---

## Step 1: Choose Your 311 Violation Type

For this lab, I focus on **Graffiti Removal** service requests in the City of Chicago. Graffiti removal is a useful 311 category for spatial predictive analysis because:

- It is a **high-volume** request type that occurs across many neighborhoods.
- It reflects both **physical conditions** (e.g., walls, alleys, commercial corridors) and **social dynamics** (e.g., disorder, enforcement, community reporting).
- The city has a clear operational interest in **prioritizing** where to send removal crews.

Studying graffiti removal requests can help the city anticipate where new incidents are likely to appear, allocate service resources more efficiently, and evaluate whether certain neighborhoods are consistently underserved.

In the rest of the analysis, I will use graffiti removal requests from **2017** as my main modeling dataset, and (optionally) use **2018** data for temporal validation.

```{r load-graffiti}
library(tidyverse)
library(lubridate)

graffiti <- read_csv("data/311_Service_Requests_-_Graffiti_Removal_-_Historical_20251115.csv")

graffiti <- graffiti %>%
  mutate(
    creation_date = mdy(`Creation Date`)  
  )

graffiti_2017 <- graffiti %>%
  mutate(year = year(creation_date)) %>%
  filter(year == 2017)

nrow(graffiti_2017)
```

---

## Step 2: Complete the Analysis

Using the class exercise as your template, adapt the code to analyze your chosen 311 violation type.

### What to Include

Work through all major sections of the analysis:

#### Part 1: Data Loading & Exploration
In this part, I load the Chicago boundary file and turn the 2017 graffiti removal records into spatial points so I can see where they are located.

```{r}

### 2.1 Load Chicago community areas
library(sf)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(spdep)
library(spatialreg)
library(MASS)

chicago_areas <- st_read("data/chicago_community_areas.shp")
```

```{r to-sf}
graffiti_sf <- graffiti_2017 %>%
  filter(!is.na(Longitude), !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"),
           crs = 4326,
           remove = FALSE)

```

```{r fix-crs}
st_crs(chicago_areas)
st_crs(graffiti_sf)

if (is.na(st_crs(chicago_areas))) {
  chicago_areas <- st_set_crs(chicago_areas, 4326)
}

if (is.na(st_crs(graffiti_sf))) {
  graffiti_sf <- st_set_crs(graffiti_sf, 4326)
}

st_crs(chicago_areas)
st_crs(graffiti_sf)

chicago_areas_proj <- st_transform(chicago_areas, 26971)
graffiti_proj      <- st_transform(graffiti_sf, 26971)
```



#### Part 2: Fishnet Grid Creation
- Create a 500m x 500m fishnet grid
- Aggregate your violations to grid cells
- Visualize the count distribution
```{r part2_fishnet}
fishnet <- st_make_grid(
  chicago_areas_proj,
  cellsize = 500,
  square   = TRUE
)

fishnet <- st_sf(
  id       = seq_along(fishnet),
  geometry = fishnet
)
```

```{r part2_counts}
idx    <- st_intersects(fishnet, graffiti_proj)
counts <- lengths(idx)

fishnet_counts <- st_sf(
  id       = seq_along(fishnet),
  count    = counts,
  geometry = fishnet$geometry
)
```

```{r part2_plot_fishnet}
library(ggplot2)

ggplot() +
  geom_sf(data = fishnet_counts, aes(fill = count), color = NA) +
  geom_sf(data = chicago_areas_proj, fill = NA, color = "white", size = 0.2) +
  labs(
    title = "Graffiti Removal Counts per 500m Grid (2017)",
    subtitle = "City of Chicago"
  ) +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal()
```


#### Part 3: Spatial Features
- Calculate k-nearest neighbor features
- Perform Local Moran's I analysis
- Identify hot spots and cold spots
- Create distance-to-hotspot measures
- Join any additional contextual data if you are looking for more to do and really get into this (e.g., demographics, land use)
```{r part3_centroids}
grid_centroids <- st_centroid(fishnet_counts)

coords <- st_coordinates(grid_centroids)
```

```{r part3_knn}
knn_nb  <- knearneigh(coords, k = 8)
nb      <- knn2nb(knn_nb)

lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

fishnet_features <- fishnet_counts %>%
  mutate(
    knn_lag_count = lag.listw(lw, count, zero.policy = TRUE)
  )
```

```{r part3_local_moran_fixed2}

localI    <- localmoran(fishnet_features$count, lw, zero.policy = TRUE)
localI_df <- as.data.frame(localI)

print(names(localI_df))

fishnet_features$Ii      <- localI_df[, 1]
fishnet_features$Z_Ii    <- localI_df[, "Z.Ii"]
fishnet_features$local_p <- localI_df[, ncol(localI_df)]  

mean_count <- mean(fishnet_features$count)

fishnet_features <- fishnet_features %>%
  mutate(
    cluster_type = case_when(
      count >= mean_count & Ii > 0 & local_p < 0.05 ~ "High-High",
      count <  mean_count & Ii > 0 & local_p < 0.05 ~ "Low-Low",
      count >= mean_count & Ii < 0 & local_p < 0.05 ~ "High-Low",
      count <  mean_count & Ii < 0 & local_p < 0.05 ~ "Low-High",
      TRUE                                          ~ "Not significant"
    )
  )

```
```{r part3_plot_clusters_fixed}
ggplot() +
  geom_sf(data = fishnet_features, aes(fill = cluster_type), color = NA) +
  geom_sf(data = chicago_areas_proj, fill = NA, color = "white", size = 0.2) +
  scale_fill_manual(
    values = c(
      "High-High"       = "red",
      "Low-Low"         = "blue",
      "High-Low"        = "orange",
      "Low-High"        = "lightblue",
      "Not significant" = "grey85"
    )
  ) +
  labs(
    title = "Local Moran's I Clusters for Graffiti Grid Counts"
  ) +
  theme_minimal()
```
```{r part3_hotspot_dist_fixed}
hotspots <- fishnet_features %>%
  filter(cluster_type == "High-High")

if(nrow(hotspots) > 0){
  hotspot_centroids <- st_centroid(hotspots)
  all_centroids     <- st_centroid(fishnet_features)

  dist_matrix <- st_distance(all_centroids, hotspot_centroids)
  fishnet_features$dist_hotspot_m <- apply(dist_matrix, 1, min)
} else {
  fishnet_features$dist_hotspot_m <- NA_real_
}
```

#### Part 4: Count Regression Models
- Fit Poisson regression
- Fit Negative Binomial regression
- Compare model fit (AIC)
```{r}
model_df <- fishnet_features |>
  st_drop_geometry() |>
  mutate(
    dist_hotspot_km = as.numeric(dist_hotspot_m) / 1000
  ) |>
  filter(
    !is.na(count),
    !is.na(knn_lag_count),
    !is.na(dist_hotspot_km)
  )
```

```{r part4_poisson}
pois_model <- glm(
  count ~ knn_lag_count + dist_hotspot_km,
  family = "poisson",
  data   = model_df
)

summary(pois_model)
pois_aic <- AIC(pois_model)
pois_aic
```
```{r part4_nb}

model_df_clean <- model_df %>%
  mutate(
    dist_hotspot_km = as.numeric(dist_hotspot_m) / 1000
  ) %>%
  dplyr::select(
    count,
    knn_lag_count,
    dist_hotspot_km
  ) %>%

  mutate(
    count          = as.numeric(count),
    knn_lag_count  = as.numeric(knn_lag_count),
    dist_hotspot_km = as.numeric(dist_hotspot_km)
  ) %>%

  filter(
    if_all(
      .cols = everything(),
      .fns  = ~ !is.na(.) & is.finite(.)
    )
  )

summary(model_df_clean)
str(model_df_clean)

glm_fit <- glm(
  count ~ knn_lag_count + dist_hotspot_km,
  data   = model_df_clean,
  family = poisson(link = "log")
)

summary(glm_fit)

coef_table <- broom::tidy(glm_fit)
coef_table

```
1. Lagged crime (knn_lag_count)

Coefficient: 0.00749

p-value: < 0.001

Interpretation:
For each additional crime in nearby cells (KNN lag), the expected crime count in the focal cell increases by about 0.75%
(because exp(0.00749) ≈ 1.0075).
This confirms strong spatial clustering: crime tends to occur near other crime.

2. Distance to hotspot (dist_hotspot_km)

Coefficient: –0.491

p-value: < 0.001

Interpretation:
For each additional kilometer of distance from the hotspot, expected crime counts decrease by 38.9%
(exp(–0.491) ≈ 0.611).
The farther from the hotspot, the fewer crimes—this is the expected spatial decay effect.

3. Model Fit

Null deviance: 455,129

Residual deviance: 97,789

This represents a major improvement (large drop), meaning the predictors explain a large share of spatial variation in crime intensity.

AIC = 106,700, which is reasonable for a grid cell Poisson model of 5,880 observations.

4. Overall conclusion

The model clearly shows:

Local crime density strongly predicts crime count (positive and significant).

Distance from hotspot strongly reduces crime (negative and significant).

Both predictors are highly significant, consistent with theories of spatial clustering.

#### Part 5: Spatial Cross-Validation (2017)
- Implement Leave-One-Group-Out cross-validation on 2017 data
- Calculate and report error metrics (MAE, RMSE)
```{r}
model_df_2017 <- model_df %>%
  filter(year == 2017) %>%
  mutate(
    dist_hotspot_km = as.numeric(dist_hotspot_m) / 1000
  ) %>%
  dplyr::select(
    count,
    knn_lag_count,
    dist_hotspot_km,
    group_id   
  ) %>%
  mutate(
    count           = as.numeric(count),
    knn_lag_count   = as.numeric(knn_lag_count),
    dist_hotspot_km = as.numeric(dist_hotspot_km)
  ) %>%
  filter(
    if_all(
      .cols = everything(),
      .fns  = ~ !is.na(.) & is.finite(.)
    )
  )

summary(model_df_2017)
str(model_df_2017)
```

#### Part 6: Model Evaluation
- Generate final predictions for both years
- Compare to KDE baseline
- Map prediction errors for 2018
- Assess model performance across time

#### CHALLENGE TASK: Temporal Validation (2018) *not mandatory, only if you want very strong spatial analytics muscles*
- Download 2018 crimes https://data.cityofchicago.org/Public-Safety/Crimes-2018/3i3m-jwuy/about_data
- Be sure to filter for BURGLARIES (FORCED ENTRY only to match the 2017 data)
- Aggregate 2018 violations to the same fishnet grid
- Use your 2017 model to predict 2018 counts
- Calculate temporal validation metrics
- Compare spatial vs. temporal validation performance

---

## Step 3: Write Your Analysis

### Critical Requirement: Explain Each Step

**For each major section, you must explain in your own words:**

- **What** you are doing in that step
In this step, I cleaned the raw crime data and constructed the spatial variables needed for the model. I converted hotspot distances from meters to kilometers and created a KNN-lagged crime count that captures activity in the surrounding grid cells. I also filtered out missing and non-finite values to ensure the regression model could run without errors.
I fit a Poisson regression model using crime counts as the dependent variable and two spatial predictors: KNN-lagged crime and distance to the nearest hotspot. The Poisson model is suitable because the outcome is a non-negative count.

- **Why** this step is important for the analysis
Preparing the data carefully is important because both Poisson regression and spatial cross-validation are sensitive to missing or inconsistent values. Spatial features such as KNN lags and distances allow the model to capture neighborhood-level effects, which are essential for understanding spatial clustering in crime.
Hotspot models provide a way to quantify how local spatial context affects crime. A positive coefficient indicates places where crime is “pulled” upward by nearby activity, while a negative coefficient indicates places where crime declines as distance increases.

- **What** you found or learned from the results
The cleaned dataset contained 5,880 observations. Crime counts were heavily right-skewed, with many zero-crime cells but a few very high-crime ones, confirming that a count model such as Poisson is appropriate.
The results show strong spatial clustering. The KNN lag variable had a positive and highly significant coefficient, meaning that crime in nearby cells increases expected crime in the focal cell. The distance variable had a negative coefficient, showing that crime decreases rapidly as distance from the hotspot increases. Together, these results confirm that both local neighbors and proximity to hotspots meaningfully shape crime intensity.

Do not simply copy text from the example or from your AI friend. 
Think about the purpose of each technique and articulate it in your own words.

---

## Step 4: Format Your Document

### Formatting Requirements

Your knitted HTML document should be **professional and easy to read**:

#### ✓ Clean Code
- Remove unnecessary code, comments, or debugging lines
- Keep only essential code chunks
- Use `code-fold: show` in your YAML header

#### ✓ Clear Structure
- Use headers to organize sections
- Include a table of contents
- Add your name and date

#### ✓ Readable Text
- Ensure all text renders properly (no broken markdown)
- Check that headers appear as headers (not plain text)
- Use proper markdown formatting

#### ✓ Quality Visualizations
- All plots should have titles and labels
- Maps should be readable
- Use consistent color schemes

#### ✓ Professional Presentation
- Proofread for typos
- Remove any "Your answer here" placeholders
- Make sure all code runs without errors
- Make IT NEAT. WE GOTTA LOOK GOOD HERE.

---

## Submission Checklist

Before you submit, verify that your document includes:

### Required Components

- [ ] **Introduction**: Brief description of your chosen 311 violation type and why you selected it
- [ ] **Data Exploration**: Maps and summary statistics of your violation data
- [ ] **Fishnet Grid**: Successfully created and visualized grid with aggregated counts
- [ ] **Spatial Features**: At least 2-3 spatial features calculated (k-NN, distance measures, etc.)
- [ ] **Local Moran's I**: Spatial autocorrelation analysis with interpretation
- [ ] **Count Regression**: Both Poisson and Negative Binomial models with comparison
- [ ] **Spatial Cross-Validation**: LOGO CV implemented on 2017 data with reported metrics
- [ ] **Temporal Validation**: 2018 data aggregated and tested with 2017 model(OPTIONAL.)
- [ ] **Validation Comparison**: Comparison of spatial vs. temporal validation performance
- [ ] **Model Comparison**: Your model vs. KDE baseline for 2018 predictions
- [ ] **Error Analysis**: Maps and discussion of where model performs well/poorly
- [ ] **Written Explanations**: Your own words explaining each step


### Technical Requirements

- [ ] Document knits to HTML without errors
- [ ] Code is clean and well-organized
- [ ] All visualizations display properly
- [ ] Headers and formatting render correctly
- [ ] File is named successfully posted to your website

---


